---
title: "Mulitomics Analysis"
author: "Kennedy Zinn"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    code_folding: hide
---
```{r}
library(Seurat)
library(hdf5r)

rna_files <- list(
  "E11" = "E11_5_filtered_gene_bc_matrices_h5.h5", 
  "E12" = "E12_5_filtered_gene_bc_matrices_h5.h5", 
  "E13" = "E13_5_filtered_gene_bc_matrices_h5.h5", 
  "E14" = "E14_5_filtered_gene_bc_matrices_h5.h5", 
  "E15" = "E15_5_filtered_gene_bc_matrices_h5.h5", 
  "E16" = "E16_5_filtered_gene_bc_matrices_h5.h5")
rna_list <- lapply(rna_files, Read10X_h5)

E11_rna <- rna_list$E11
E12_rna <- rna_list$E12
E13_rna <- rna_list$E13
E14_rna <- rna_list$E14
E15_rna <- rna_list$E15
E16_rna <- rna_list$E16

remove(rna_list)
```

# Repeated Measures Analysis

## Visualize Trajectory by Cell Type

```{r, include = FALSE}
library(SingleR)

E11_seurat <- CreateSeuratObject(E11_rna, project = "E11.5")
E12_seurat <- CreateSeuratObject(E12_rna, project = "E12.5")
E13_seurat <- CreateSeuratObject(E13_rna, project = "E13.5")
E14_seurat <- CreateSeuratObject(E14_rna, project = "E14.5")
E15_seurat <- CreateSeuratObject(E15_rna, project = "E15.5")
E16_seurat <- CreateSeuratObject(E16_rna, project = "E16.5")

combined <- merge(E11_seurat, y = c(E12_seurat, E13_seurat, E14_seurat, E15_seurat, E16_seurat),
                  add.cell.ids = c("E11.5", "E12.5", "E13.5", "E14.5", "E15.5", "E16.5"))
remove(list = ls(pattern = "^E"))
```

```{r}
combined$mitoPercent <- PercentageFeatureSet(combined, pattern = "^mt-")
combined <- subset(combined, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 &
                     nCount_RNA > 800 &
                     mitoPercent < 5)
```


```{r}
# preprocessing
combined <- NormalizeData(combined)
combined <- FindVariableFeatures(combined)
combined <- ScaleData(combined)
combined <- RunPCA(combined)

# integrate
library(harmony)
combined <- RunHarmony(combined, group.by.vars = "orig.ident")
combined <- RunUMAP(combined, reduction = "harmony", dims = 1:30)

# clustering using KNN
combined <- FindNeighbors(combined, reduction = "harmony", dims = 1:30)
combined <- FindClusters(combined, resolution = 0.1)
```

```{r}
# get annotation
library(celldex)
ref <- MouseRNAseqData() 

# annotate
library(SingleCellExperiment)
combined <- JoinLayers(combined)
sce <- as.SingleCellExperiment(combined)

predictions <- SingleR(test = sce, ref = ref, labels = ref$label.fine)
combined$cell_type <- predictions$labels
remove(sce)
```

From these plots, we can see some clustering by cell type and time point. There is some overlap in time points and the clusters identified by PCA and UMAP, suggesting that other variables not recorded in this dataset contribute to neurogenesis. (such as neuron type/ fate)

```{r}
# clusters found by UMAP and PCA
p <- DimPlot(combined, reduction = "umap", label = TRUE)
p$layers[[1]]$aes_params$alpha <- 0.25
p

# by timepoint
p <- DimPlot(combined, reduction = "umap", group.by = "orig.ident", label = F)
p$layers[[1]]$aes_params$alpha <- 0.25
p

# by cell type

p <- DimPlot(combined, reduction = "umap", group.by = "cell_type", label = F, repel = TRUE)
p$layers[[1]]$aes_params$alpha <- 0.25
p
```

```{r}
# Generate distinct colors
library(scales)
cell_types <- unique(combined$cell_type)
n_colors <- length(cell_types)
my_colors <- hue_pal()(n_colors) 

# Manually adjust colors for similar cell types
# Example: if "Neurons_1" and "Neurons_2" are too similar
names(my_colors) <- cell_types
my_colors["Neurons"] <- "#A020F0"  # Red
my_colors["NPCs"] <- "#377EB8"  # Blue
my_colors["OPCs"] <- "#E41A1C"   # Green
my_colors["aNSCs"] <- "#4DAF4A" # purple
p <- DimPlot(combined, reduction = "umap", group.by = "cell_type", cols = my_colors, label = F, repel = TRUE)
p$layers[[1]]$aes_params$alpha <- 0.5
p
```

```{r, include = FALSE}
library(dplyr)
# add corticogenic wave labels
combined$corticogenic_wave <- case_when(
  combined$orig.ident %in% c("E11.5", "E12.5", "E13.5") ~ "Deep Layers (VI, V)",
  combined$orig.ident %in% c("E14.5", "E15.5", "E16.5") ~ "Superficial Layers (II, III, IV)",
  TRUE ~ "Other"
)

# Or using base R:
combined$corticogenic_wave <- ifelse(
  combined$orig.ident %in% c("E11.5", "E12.5", "E13.5"),
  "Deep Layers (VI, V)",
    "Superficial Layers (II, III, IV)"
)
```

UMAP by corticogenesis, using time points by Di Bella et al.(2021) and Ohte et al. (2025)
```{r}
# UMAP by layer
library(ggplot2)
p <- DimPlot(combined, 
        reduction = "umap", 
        group.by = "corticogenic_wave",
        cols = c("Deep Layers (VI, V)" = "#E41A1C",
                 "Superficial Layers (II, III, IV)" = "#377EB8")) +
  ggtitle("Deep vs Superficial Layers")
p$layers[[1]]$aes_params$alpha <- 0.25
p
```

# Temporal Analysis of Corticogenesis

## Differential Analysis of Neural Progenitor Cells

```{r}
# Compare consecutive timepoints to see temporal changes

# set identities to timepoints
combined$ct.time <- paste0(combined$cell_type, combined$orig.ident)

Idents(combined) <- "ct.time"

contrasts_list <- list(
  E12_vs_E11 = c("NPCsE12.5", "NPCsE11.5"),
  E13_vs_E12 = c("NPCsE13.5", "NPCsE12.5"),
  E14_vs_E13 = c("NPCsE14.5", "NPCsE13.5"),
  E15_vs_E14 = c("NPCsE15.5", "NPCsE14.5"),
  E16_vs_E15 = c("NPCsE16.5", "NPCsE15.5")
)

# Run all contrasts
contrast_results <- lapply(names(contrasts_list), function(contrast_name) {
  contrast <- contrasts_list[[contrast_name]]
  
  de_genes <- FindMarkers(combined,
                          ident.1 = contrast[1],
                          ident.2 = contrast[2],
                          min.pct = .1,
                          logfc.threshold = 0)
  
  de_genes$gene <- rownames(de_genes)
  de_genes$contrast <- contrast_name
  return(de_genes)
})

# Combine all results
all_contrasts <- do.call(rbind, contrast_results)

# Summary
contrast_summary <- all_contrasts %>%
  filter(p_val_adj < 0.05) %>%
  group_by(contrast) %>%
  summarise(
    n_up = sum(avg_log2FC > 0),
    n_down = sum(avg_log2FC < 0),
    total_de = n()
  )

print(contrast_summary)
```

### Bar plot of DEG Counts 

```{r}
library(tidyr)
# Summary of significant genes per contrast
de_summary <- all_contrasts %>%
  filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.5) %>%
  group_by(contrast) %>%
  summarise(
    Upregulated = sum(avg_log2FC > 0),
    Downregulated = sum(avg_log2FC < 0)
  ) %>%
  pivot_longer(cols = c(Upregulated, Downregulated),
               names_to = "Direction",
               values_to = "Count")

# Order contrasts chronologically
de_summary$contrast <- factor(de_summary$contrast,
                              levels = c("E12_vs_E11", "E13_vs_E12", "E14_vs_E13", 
                                        "E15_vs_E14", "E16_vs_E15"))

ggplot(de_summary, aes(x = contrast, y = Count, fill = Direction)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Upregulated" = "#E41A1C", "Downregulated" = "#377EB8")) +
  theme_minimal() +
  labs(title = "Number of Differentially Expressed Genes Per Timepoint Comparison",
       x = "Contrast",
       y = "Number of DE Genes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Volcano Plots

```{r}
library(patchwork)

# Create volcano plot for each contrast
names(contrast_results) <- names(contrasts_list)

volcano_plots <- lapply(names(contrast_results), function(contrast_name) {
  data <- contrast_results[[contrast_name]]
  data$significant <- ifelse(data$p_val_adj < 0.05 & abs(data$avg_log2FC) > 0.5, 
                             "Significant", "Not Significant")
  
  ggplot(data, aes(x = avg_log2FC, y = -log10(p_val_adj), color = significant)) +
    geom_point(alpha = 0.6, size = 1) +
    scale_color_manual(values = c("Not Significant" = "gray", "Significant" = "red")) +
    geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "blue") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
    theme_minimal() +
    labs(title = contrast_name,
         x = "Log2 Fold Change",
         y = "-Log10 Adjusted P-value") +
    theme(legend.position = "bottom")
})

# Combine all volcano plots
volcano_plots[[1]] + volcano_plots[[2]] + volcano_plots[[3]] + 
  volcano_plots[[4]] + volcano_plots[[5]] + 
  plot_layout(ncol = 3)
```

# Gene Profile
```{r}
# identify layer-specific DEGs
all_contrasts$cort_layer <- ifelse(grepl("^E12|^E13", all_contrasts$contrast),
                                 "deep",
                                 "superficial")
top30 <- all_contrasts %>% 
  filter(p_val_adj < 0.05) %>% 
  group_by(contrast) %>% 
  arrange(-abs(avg_log2FC), .by_group = T) %>% 
  slice_head(n = 30) %>% 
  ungroup
```

```{r}
deep_top <- top30[top30$cort_layer=="deep",] %>% 
  arrange(-abs(avg_log2FC))
deep_top
```

```{r}
sup_top <- top30[top30$cort_layer=="superficial",] %>% 
  arrange(-abs(avg_log2FC))
sup_top
```

```{r}
#DEGs unique to deep layer
deep_unique <- anti_join(deep_top, sup_top, by = "gene")
deep_unique
```

```{r}
sup_unique <- anti_join(sup_top, deep_top, by = "gene")
sup_unique
```


```{r}

library(dplyr)
library(ggplot2)

plot_gene_by_time <- function(seurat_obj, gene, time_col = "orig.ident") {
  
  df <- FetchData(seurat_obj, vars = c(gene, time_col)) %>%
    group_by(.data[[time_col]]) %>%
    summarise(avg_expr = mean(.data[[gene]]), .groups = "drop")
  
  ggplot(df, aes_string(x = time_col, y = "avg_expr", group = 1)) +
    geom_line(size = 1, color = "steelblue") +
    geom_point(size = 3, color = "darkblue") +
    labs(
      title = paste("Average expression of", gene, "across time points"),
      x = "Time point",
      y = "Average expression"
    ) +
    theme_bw()
}
```

Xist is exclusively expressed on the inactive X chromosome and is essential for X-inactivation (NIH). This gene was the most differentially expressed in both groups, as its expression changed dramatically from E12.5 to E14.5.

```{r}
# plot Xist expression across neurogenesis
plot_gene_by_time(combined, "Xist")
```

Kcnn2 enables small conductance calcium-activate potassium channel activity. It is used for modulating chemical synaptic transmission and is active in glutamatergic synapses and postsynaptic membrane (NIH). This gene was a unique top DEG in early neurogensis. However, but unsurprisingly, the plot shows that expression increases throughout neurogenesis, peaking at E14.5 and leveling out after; the most significant increase in expression was detected in the early phases, from E11.5 to E12.5.

```{r}
plot_gene_by_time(combined, "Kcnn2")
```

Hbb-bs top unique DEG in the later phases of neurogenesis. Hbb-bs code for a polyeptide chain found in hemoglobin and is involved in transporting oxygen to peripheral tissues. Peaking in expression at E12.5 and dropping by E13.5, Hbb-bs expression gradually increases linearly from E13.5 to E16.5. Literature suggests that neuronal hemoglobin expression in mature neurons may play a role in neuronal epigenetics, metabolism and function (Brown et al., 2016, Richter et al., 2009).

```{r}
plot_gene_by_time(combined, "Hbb-bs")
```

# ATAC-Seq; Multiomics Integration

```{r, include=FALSE}
library(Signac)
rm(list = ls())
E18_atac <- Read10X_h5("E18_5_filtered_peak_bc_matrix.h5")
E18_frag_file <- "E18_atac_fragments.tsv.tar.download/E18_atac_fragments/E18_atac_fragments.tsv.gz"
```

```{r, include = FALSE}
# Get annotations from EnsDb
library(EnsDb.Mmusculus.v79)
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Mmusculus.v79)

# Change to UCSC style
seqlevelsStyle(annotations) <- "UCSC"
genome(annotations) <- "mm10"
```

```{r}
# Same for E18.5
E18_chromatin <- CreateChromatinAssay(
  counts = E18_atac,
  sep = c(":", "-"),
  fragments = E18_frag_file,
  genome = "mm10",
  min.cells = 10,
  min.features = 200,
  annotation = annotations
)

E18_seurat <- CreateSeuratObject(
  counts = E18_chromatin,
  assay = "ATAC",
  project = "E18.5_ATAC"
)
```

```{r}
# processing
E18_seurat <- RunTFIDF(E18_seurat, assay = "ATAC")
E18_seurat <- FindTopFeatures(E18_seurat, min.cutoff = 'q0')
E18_seurat <- RunSVD(E18_seurat, assay = "ATAC")
E18_seurat <- RunUMAP(E18_seurat, reduction = "lsi", dims = 2:30)
E18_seurat <- FindNeighbors(E18_seurat, reduction = "lsi", dims = 2:30)
E18_seurat <- FindClusters(E18_seurat, resolution = 0.5, algorithm = 3)

# visualize
DimPlot(E18_seurat, reduction = "umap", label = TRUE)
```


```{r}
gene.activities <- GeneActivity(E18_seurat)

E18_seurat[['RNA']] <- CreateAssayObject(counts = gene.activities)
E18_seurat <- NormalizeData(object = E18_seurat,
              assay = 'RNA',
              normalization.method = 'LogNormalize',
              scale.factor = median(E18_seurat$nCount_RNA))

DefaultAssay(E18_seurat) <- 'RNA'

FeaturePlot(E18_seurat, features = c('Pax6', 'Sox2', 'Dcx', 'Dlx1'), # canonical marker genes of progenitor neurons
            pt.size = 0.1,
            max.cutoff = 'q95',
            ncol = 3)
```

```{r}
# Load the scRNA-seq data
E15_rna <- Read10X_h5("E15_5_filtered_gene_bc_matrices_h5.h5")
E15_rna <- CreateSeuratObject(E15_rna, project = "E15.5")

E15_rna$mitoPercent <- PercentageFeatureSet(E15_rna, pattern = "^mt-")
E15_rna <- subset(E15_rna, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 &
                     nCount_RNA > 800 &
                     mitoPercent < 5)
```


```{r, include=FALSE}
# preprocessing
E15_rna <- NormalizeData(E15_rna)
E15_rna <- FindVariableFeatures(E15_rna)
E15_rna <- ScaleData(E15_rna)
E15_rna <- RunPCA(E15_rna)
E15_rna <- RunUMAP(E15_rna, dims = 1:30)

# clustering using KNN
E15_rna <- FindNeighbors(E15_rna, dims = 1:30)
E15_rna <- FindClusters(E15_rna, resolution = 0.1)
```

```{r}
# get annotation
library(celldex)
ref <- MouseRNAseqData() 

# annotate
library(SingleCellExperiment)
E15_rna <- JoinLayers(E15_rna)
sce <- as.SingleCellExperiment(E15_rna)

predictions <- SingleR(test = sce, ref = ref, labels = ref$label.fine)
E15_rna$cell_type <- predictions$labels
remove(sce)
```

```{r}
# identify anchors
transfer.anchors <- FindTransferAnchors(reference = E15_rna,
                    query = E18_seurat,
                    reduction = 'pcaproject')


predicted.labels <- TransferData(anchorset = transfer.anchors,
             refdata = E15_rna$cell_type,
             weight.reduction = E18_seurat[['lsi']],
             dims = 2:30)
head(predicted.labels)

E18_seurat <- AddMetaData(object = E18_seurat, metadata = predicted.labels)


plot1 <- DimPlot(E18_seurat, 
        reduction = 'umap',
        group.by = 'predicted.id',
        label = TRUE,
        repel = TRUE) + NoLegend() + ggtitle('scATAC-Seq')

plot2 <- DimPlot(E15_rna, 
                 reduction = 'umap',
                 group.by = 'cell_type',
                 label = TRUE,
                 repel = TRUE) + NoLegend() + ggtitle('scRNA-Seq')
plot1
plot2
```

## Predict Chromatic Accessibility from Gene Expression

```{r}
# Impute gene expression values to ATAC cells
genes.use <- VariableFeatures(E15_rna)

# Transfer gene expression data to ATAC cells
refdata <- GetAssayData(E15_rna, assay = "RNA", slot = "data")[genes.use, ]

imputation <- TransferData(
  anchorset = transfer.anchors,
  refdata = refdata,
  weight.reduction = E18_seurat[['lsi']],
  dims = 2:30
)

# Add imputed RNA data to ATAC object
E18_seurat[["imputed_RNA"]] <- imputation

# Compare gene activity (from ATAC) vs imputed gene expression (from RNA)
DefaultAssay(E18_seurat) <- "imputed_RNA"
plot_imputed <- FeaturePlot(E18_seurat, 
                            features = c('Pax6', 'Sox2', 'Dcx', 'Dlx1'),
                            pt.size = 0.1,
                            max.cutoff = 'q95',
                            ncol = 2) + 
  plot_annotation(title = 'Imputed Gene Expression from scRNA-seq')

DefaultAssay(E18_seurat) <- "RNA"
plot_activity <- FeaturePlot(E18_seurat, 
                             features = c('Pax6', 'Sox2', 'Dcx', 'Dlx1'),
                             pt.size = 0.1,
                             max.cutoff = 'q95',
                             ncol = 2) + 
  plot_annotation(title = 'Gene Activity from scATAC-seq')

# Side-by-side comparison
plot_imputed / plot_activity
```

```{r}
# Extract data for correlation analysis
genes_to_test <- c('Pax6', 'Sox2', 'Dcx', 'Dlx1', 'Eomes', 'Tbr1', 'Neurod1')
genes_to_test <- genes_to_test[genes_to_test %in% rownames(E18_seurat[["RNA"]])]

correlation_results <- data.frame()

for (gene in genes_to_test) {
  # Gene activity from ATAC
  activity <- E18_seurat[["RNA"]]@data[gene, ]
  
  # Imputed expression from RNA
  expression <- E18_seurat[["imputed_RNA"]]@data[gene, ]
  
  # Calculate correlation
  cor_result <- cor.test(activity, expression, method = "spearman")
  
  correlation_results <- rbind(correlation_results, data.frame(
    gene = gene,
    correlation = cor_result$estimate,
    p_value = cor_result$p.value
  ))
}

print("Correlation between Gene Activity (ATAC) and Imputed Expression (RNA):")
print(correlation_results)

# Visualize correlations
library(ggplot2)
ggplot(correlation_results, aes(x = reorder(gene, correlation), y = correlation)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(correlation, 2)), vjust = -0.5) +
  labs(title = "Gene Expression vs Chromatin Accessibility Correlation",
       x = "Gene", y = "Spearman Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
# Link ATAC peaks to genes based on correlation
DefaultAssay(E18_seurat) <- "ATAC"

# This requires the imputed RNA data
library(BSgenome.Mmusculus.UCSC.mm10)
E18_seurat <- RegionStats(
  object = E18_seurat,
  genome = BSgenome.Mmusculus.UCSC.mm10
)

E18_seurat <- LinkPeaks(
  object = E18_seurat,
  peak.assay = "ATAC",
  expression.assay = "imputed_RNA",
  genes.use = genes_to_test
)

# Visualize peak-gene links
CoveragePlot(
  object = E18_seurat,
  region = "Pax6",
  features = "Pax6",
  expression.assay = "imputed_RNA",
  extend.upstream = 50000,
  extend.downstream = 50000
)
```

```{r}
# identify potential predictors of Pax6 chromatin accessibility

pax6_accessibility <- E18_seurat[["RNA"]]@data["Pax6", ]

# Get all genes from imputed RNA (predictor variables)
genes_to_test <- rownames(E18_seurat[["imputed_RNA"]])

cat(paste("Testing correlation of", length(genes_to_test), "genes with Pax6 chromatin accessibility...\n"))

correlation_results <- data.frame()

# Loop through all genes to find which predict Pax6 accessibility
for (gene in genes_to_test) {
  # Imputed expression of each gene from E15.5 RNA
  expression <- E18_seurat[["imputed_RNA"]]@data[gene, ]
  
  # Skip genes with no variation
  if (sd(expression) == 0) next
  
  # Calculate correlation
  cor_result <- cor.test(expression, pax6_accessibility, method = "spearman")
  
  correlation_results <- rbind(correlation_results, data.frame(
    gene = gene,
    correlation = cor_result$estimate,
    p_value = cor_result$p.value
  ))
}

# Adjust p-values for multiple testing
correlation_results$p_adjusted <- p.adjust(correlation_results$p_value, method = "BH")

# Sort by absolute correlation strength
correlation_results <- correlation_results[order(abs(correlation_results$correlation), decreasing = TRUE), ]

# Display top results
cat("\nTop 20 genes whose expression correlates with Pax6 chromatin accessibility:\n")
print(head(correlation_results, 20))

# Get significant genes (adjusted p-value < 0.05)
significant_genes <- correlation_results[correlation_results$p_adjusted < 0.05, ]
cat(paste("\nNumber of significant genes (FDR < 0.05):", nrow(significant_genes), "\n"))

# Visualize top 30 correlations
top_genes <- head(correlation_results, 30)

library(ggplot2)

# Separate visualization for positive and negative correlations
pos_genes <- head(correlation_results[correlation_results$correlation > 0, ], 15)
neg_genes <- head(correlation_results[correlation_results$correlation < 0, ], 15)

p2 <- ggplot(pos_genes, aes(x = reorder(gene, correlation), y = correlation)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(correlation, 2)), hjust = -0.1, size = 3) +
  labs(title = "Top Positive Predictors of Pax6 Accessibility",
       x = "Gene", y = "Spearman Correlation") +
  coord_flip() +
  theme_minimal()

p3 <- ggplot(neg_genes, aes(x = reorder(gene, correlation), y = correlation)) +
  geom_bar(stat = "identity", fill = "coral") +
  geom_text(aes(label = round(correlation, 2)), hjust = 1.1, size = 3) +
  labs(title = "Top Negative Predictors of Pax6 Accessibility",
       x = "Gene", y = "Spearman Correlation") +
  coord_flip() +
  theme_minimal()

library(patchwork)
p2 / p3

# Scatter plots for top predictors
top_5_genes <- head(correlation_results, 5)$gene

scatter_plots <- list()
for (i in 1:min(5, length(top_5_genes))) {
  gene <- top_5_genes[i]
  plot_data <- data.frame(
    expression = E18_seurat[["imputed_RNA"]]@data[gene, ],
    accessibility = pax6_accessibility,
    cell_type = E18_seurat$predicted.id
  )
  
  p <- ggplot(plot_data, aes(x = expression, y = accessibility, color = cell_type)) +
    geom_point(alpha = 0.5, size = 1) +
    geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
    labs(title = paste(gene, "→ Pax6 Accessibility"),
         x = paste(gene, "Expression (E15.5)"),
         y = "Pax6 Accessibility (E18.5)") +
    theme_minimal() +
    theme(legend.position = "none")
  
  scatter_plots[[i]] <- p
}

# Arrange scatter plots
wrap_plots(scatter_plots, ncol = 2)

```


# References

Brown N, Alkhayer K, Clements R, Singhal N, Gregory R, Azzam S, Li S, Freeman E, McDonough J. Neuronal Hemoglobin Expression and Its Relevance to Multiple Sclerosis Neuropathology. J Mol Neurosci. 2016 May;59(1):1-17. doi: 10.1007/s12031-015-0711-6. Epub 2016 Jan 25. PMID: 26809286; PMCID: PMC4851882.

Di Bella DJ, Habibi E, Stickels RR, Scalia G, Brown J, Yadollahpour P, Yang SM, Abbate C, Biancalani T, Macosko EZ, Chen F, Regev A, Arlotta P. Molecular logic of cellular diversification in the mouse cerebral cortex. Nature. 2021 Jul;595(7868):554-559. doi: 10.1038/s41586-021-03670-5. Epub 2021 Jun 23. Erratum in: Nature. 2021 Aug;596(7873):E11. doi: 10.1038/s41586-021-03797-5. PMID: 34163074; PMCID: PMC9006333.

Ohte N, Kimura T, Sekine R, Yoshizawa S, Furusho Y, Sato D, Nishiyama C, Hanashima C. Differential neurogenic patterns underlie the formation of primary and secondary areas in the developing somatosensory cortex. Cereb Cortex. 2025 Feb 5;35(2):bhae491. doi: 10.1093/cercor/bhae491. PMID: 39756431; PMCID: PMC11795310.

Richter F, Meurers BH, Zhu C, Medvedeva VP, Chesselet MF. Neurons express hemoglobin alpha- and beta-chains in rat and human brains. J Comp Neurol. 2009 Aug 10;515(5):538-47. doi: 10.1002/cne.22062. PMID: 19479992; PMCID: PMC3123135.
